Expert-In-The-Loop Causal Discovery: \\ Iterative Model Refinement Using Expert Knowledge

### 2025-02-10 paper/main.tex
done with intro

\section{Introduction}
Understanding cause-and-effect relationships between variables is a fundamental
objective in many scientific fields. These relationships reveal the mechanisms
behind observed phenomena and guide effective interventions or policy
decisions. Causal discovery methods aim to discover such relationships among
random variables using observational data. In the DAG literature, the primary
focus has been on developing automated algorithms to learn causal structures
from datasets. These efforts have led to numerous causal discovery algorithms,
such as constraint-based methods like PC algorithm \citep{Spirtes2001} and Fast
Causal Inference \citep{Spirtes2000}), score-based methods such as Hill-Climb
Search and Greedy Equivalence Search \citep{Chickering2002}, and continuous
optimization-based methods like NO TEARS \citep{Zheng2018} and DAGMA
\citep{Bello2022}. Despite significant progress in automated causal discovery,
their adoption in applied research has been limited. Researchers often prefer
to manually construct DAGs based on their domain expertise \citep{Tennant2020,
Petersen2021}. We attribute this preference to several challenges with existing
causal discovery algorithms in practical settings:

\begin{enumerate}
	\item \textbf{Lack of Trust:} While most algorithms are asymptotically
		consistent, their behavior on finite samples is not well
		understood. Their output can vary significantly depending on
		the choice of algorithm and hyperparameters, making it
		difficult to assess reliability. Additionally, the absence of
		robust performance evaluation methods for any given dataset
		further reduce the confidence in their outputs. 
	\item \textbf{Outputs Markov Equivalence Class (MEC):} As multiple
		DAGs can be faithful to a given observational dataset, automated 
		algorithms can only recover the MECs. These MECs can contain a
		combination of directed and undirected edges. However, most
		methods for downstream tasks, such as identification or causal
		effect estimation, assume knowledge of a fully oriented DAG. 
\end{enumerate}

Figure~\ref{fig:intro} highlights some of these issues. On the other hand,
constructing DAG based solely on domain expertise can be error prone as it
requires us to distinguish between different causal structures, such as direct
or indirect effects or causal mediation. Given these complexities, it is
important to validate the consistency of a DAG against our dataset. One way to
test this consistency is by testing whether the CI statements implied by the
DAG hold in the data \citep{Ankan2021}. Specifically, each missing edge between
a pair of variables in the DAG implies a CI statement, which can be tested in
the dataset using a statistical test. If we find any violation to CI
statements, we can make modifications to our model accordingly to fix the
violations.

@@

    \caption{A comparison of Markov Equivalence Classes (MECs) learned from the Adult
	     Income Dataset \citep{Becker1996} using different causal discovery
	     algorithms and sample sizes. Edge colors represents the sample size
	     used: red for $N=400$, and blue for $N=800$. (a) PC algorithm with a
	     mutual information based CI test, (b) PC algorithm with a
    	     residualization based test \citep{Ankan2023}, (c) Hill Climb Search with
	     Bayesian Information Criterion (BIC) score. The learned model structure varies
             significantly across different algorithms and sample sizes.}
    \label{fig:intro}
\end{figure}

In this paper, we propose a structure learning method that leverages this CI
testing approach to assist researchers during manual DAG construction, rather
than just validating the model. Our approach follows an iterative
model-building strategy that integrates domain expertise with CI testing on
data. At each iteration, we use CI testing to identify pairs of variables where
an edge is either missing or superfluous. We can then use this information
along with domain expertise to determines whether to add or remove these edges.
The orientation of the edge when adding is determined based on the domain
expertise. As at any iteration, multiple changes to the DAG are possible. We
also give a way to rank potential new edges additions to the DAG using effect
size measures for CI tests to address the most severe violations of CI tests
first. As a result this approach acts like a guide to the expert when manually
constructing their model while ensuring that the model is consistent with the
data.

Our main contributions in this paper are as follows:

\begin{enumerate}
    \item We develop an iterative method based on CI testing to assist researchers in 
	    constructing DAGs using their domain expertise (Section~\ref{sec:modification}).
    \item We show a ranking method to prioritize potential modifications to the DAG that 
    	    help in fixing the most severe violations first (Section~\ref{sec:ranking}).
    \item We show empirically that even experts that make mistakes can outperform purely data-driven
	causal discovery algorithms (Section~\ref{sec:empirical}).
    \item We provide two proof-of-concept implementations of our approach: one geared towards
	LLMs as experts, and a graphical user interface designed 
	for human experts (Section~\ref{sec:web}).
\end{enumerate}

>>>>

\section{Introduction}
Understanding cause-and-effect relationships between variables is a fundamental
objective in many scientific fields. These relationships reveal the mechanisms
behind observed phenomena and guide effective interventions or policy
decisions. Causal discovery methods aim to discover such relationships among
random variables using observational data. In the DAG literature, the primary
focus has been on developing automated algorithms to learn causal structures
from datasets. These efforts have led to numerous causal discovery algorithms,
such as constraint-based methods like PC algorithm \citep{Spirtes2001,KalischB07} 
and Fast Causal Inference \citep{Spirtes2000}), score-based methods such as Hill-Climb
Search and Greedy Equivalence Search \citep{Chickering2002}, and continuous
optimization-based methods like NOTEARS \citep{Zheng2018} and DAGMA
\citep{Bello2022}. Despite significant progress in automated causal discovery,
their adoption in applied research has been limited. Challenges encountered with existing
causal discovery algorithms in practice include but are not limited to:

\begin{enumerate}
	\item \textbf{Lack of Trust:} While constraint-based algorithms are 
		often asymptotically
		consistent \citep{KalischB07}, they can and do make mistakes
		on  finite samples. These mistakes can be severe and contradict
		obvious domain knowledge (think of edges going into unmodifiable
		attributes such as Age). The choice of algorithm and hyperparameters
		significantly aa=ffect the output, making it
		difficult to assess reliability. Additionally, the absence of
		robust performance evaluation methods for any given dataset
		further reduce the confidence in their outputs. A recent paper
		therefore advised Epidemiologists to not attempt using structure
		learning algorithms without help of an expert \citep{Gururaghavendran_2024}.
	\item \textbf{Outputs Markov Equivalence Class (MEC):} As multiple
		DAGs can be faithful to a given observational dataset, automated 
		algorithms can only recover the MECs. These MECs can contain a
		combination of directed and undirected edges. This structural uncertainty
		can make difficult or impossible to apply the learned model for downstream
		tasks, such as identification or causal
		effect estimation \citep{Maathuis_2009,PerkovicTKM17}.
\end{enumerate}

Figure~\ref{fig:intro} highlights some of these issues. In practice, DAGs 
are still largely constructed from domain knowledge alone 
\citep{Tennant2020,Petersen2021}. This can, however, be equally problematic.
Constructing DAGs requires us to distinguish between different causal structures, 
such as direct or indirect effects or causal mediation. Specifically, there is often a lack
of theoretical support for the \emph{absence} of direct causal effects, which are 
the key assumptions that graphical models make to enable downstream causal inferences.
Given the likelihood of making mistakes, it is therefore
important to at least validate the consistency of a DAG against our dataset. One way to
test this consistency is by testing whether the conditional independence 
(CI) statements implied by the DAG hold in the data \citep{Ankan2021}. Specifically, 
each missing edge between
a pair of variables in the DAG leads to one or more CI statements, 
which can be checked using statistical tests. Violations to CI
statements can point to the erroneous omission of important causal effects from a model
or to the presence of latent confounders.

@@

    \caption{A comparison of Markov Equivalence Classes (MECs) learned from the Adult
	     Income Dataset \citep{Becker1996} using different causal discovery
	     algorithms and sample sizes. Edge colors represent the sample size
	     used: red for $N=400$, and blue for $N=800$. (a) PC algorithm with a
	     mutual information based CI test, (b) PC algorithm with a
    	     residualization-based test \citep{Ankan2023}, (c) Hill Climb Search with
	     Bayesian Information Criterion (BIC) score. The learned model structure varies
             significantly across different algorithms and sample sizes.}
    \label{fig:intro}
\end{figure}

In this paper, we propose a structure learning method that leverages this CI
testing approach to assist researchers during manual DAG construction, rather
than just validating the model. Our approach follows an iterative
model-building strategy that integrates domain expertise with CI testing on
data. At each iteration, we use CI testing to identify pairs of variables that
lack (direct or indirect) connections or are connected by a superfluous edge.
 We then use this information
along with domain expertise to determine which changes.
The orientation of the edge when adding is determined based on the domain
expertise; critically, this does not require the domain expert to distinguish 
between direct and indirect effects. Our resulting approach can be 
thought of as a guidance during manual model construction that ensures the 
model remains consistent to the data.

Our contributions are organized as follows:

\begin{enumerate}
    \item In Section~\ref{sec:modification}, we present our iterative method for DAG construction
	based on domain knowledge and data, and prove its correctness in an ``oracle setting''.
    \item We then present a ranking method to prioritize potential modifications to the DAG that 
    	    help in fixing the most severe violations first (Section~\ref{sec:ranking}).
    \item We show empirically that even experts that make mistakes can outperform purely 
	data-driven
	causal discovery algorithms (Section~\ref{sec:empirical}).
    \item We provide two proof-of-concept implementations of our approach: one geared towards
	LLMs as experts, and a graphical user interface designed 
	for human experts (Section~\ref{sec:web}).
\end{enumerate}
